Mathematical Models of Delta-Notch Signalling

### 2025-03-14 paper/paper.tex
paper.tex

\subsection{Kolmogorov Forward Equations}

@@

TBD

>>>>

\subsection{Kolmogorov Forward Equations}
\subsubsection{Derivation of Kolmogorov Forward Equation}

Let us define:\\
$P(n,d,i,t)$ = probability of having $n$ Notch receptors, $d$ Delta ligands, and $i$ NICD molecules at time $t$.\\

The Kolmogorov forward equation describes how $P(n,d,i,t)$ changes over a small time step $\Delta t$:

\begin{align*}
P(n,d,i,t+\Delta t) &= P(n,d,i,t) \\
&+ P(n-1,d,i,t) \cdot H^+(i) \Delta t \quad \text{(Notch production)} \\
&+ P(n,d-1,i,t) \cdot H^-(i) \Delta t \quad \text{(Delta production)} \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1)\Delta t \quad \text{(Notch decay)} \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1)\Delta t \quad \text{(Delta decay)} \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1)\Delta t \quad \text{(NICD decay)} \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext}\Delta t \quad \text{(Trans-activation)} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]\Delta t
\end{align*}

Rearranging to find the time derivative:

\begin{align*}
\frac{P(n,d,i,t+\Delta t) - P(n,d,i,t)}{\Delta t} &= P(n-1,d,i,t) \cdot H^+(i) \\
&+ P(n,d-1,i,t) \cdot H^-(i) \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
\end{align*}

Taking the limit as $\Delta t \rightarrow 0$:

\begin{align*}
\frac{\partial P(n,d,i,t)}{\partial t} &= P(n-1,d,i,t) \cdot H^+(i) \\
&+ P(n,d-1,i,t) \cdot H^-(i) \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
\end{align*}

\subsubsection{Matrix Formulation}
The Kolmogorov forward equation can be expressed in matrix form as:
\[
\frac{d\vec{P}(t)}{dt} = \mathbf{A} \vec{P}(t)
\]

where $\vec{P}(t)$ is a column vector containing the probabilities for all possible states, and $\mathbf{A}$ is the transition rate matrix. \\

To construct the matrix $\mathbf{A}$, we first establish a one-to-one mapping between the three-dimensional state space $(n,d,i)$ and a one-dimensional index $k$. For a system with maximum values $n_{max}$, $d_{max}$, and $i_{max}$, we define:
\[
k(n,d,i) = n \cdot (d_{max}+1) \cdot (i_{max}+1) + d \cdot (i_{max}+1) + i
\]
for $0 \leq n \leq n_{max}, \; 0 \leq d \leq d_{max}, \; 0 \leq i \leq i_{max}$.

The inverse mapping gives:
\begin{align*}
n(k) &= \left\lfloor \frac{k}{(d_{max}+1) \cdot (i_{max}+1)} \right\rfloor \\
d(k) &= \left\lfloor \frac{k \bmod ((d_{max}+1) \cdot (i_{max}+1))}{i_{max}+1} \right\rfloor \\
i(k) &= k \bmod (i_{max}+1)
\end{align*}

With this mapping, we can now define the elements of matrix $\mathbf{A}$. Let $k$ and $k'$ be the indices corresponding to states $(n,d,i)$ and $(n',d',i')$ respectively:

\begin{align*}
A_{k',k} = 
\begin{cases}
H^+(i) & \text{if } (n',d',i') = (n+1,d,i) \quad \text{(Notch production)} \\
H^-(i) & \text{if } (n',d',i') = (n,d+1,i) \quad \text{(Delta production)} \\
\gamma(n) & \text{if } (n',d',i') = (n-1,d,i) \quad \text{(Notch decay)} \\
\gamma(d) & \text{if } (n',d',i') = (n,d-1,i) \quad \text{(Delta decay)} \\
\gamma_I(i) & \text{if } (n',d',i') = (n,d,i-1) \quad \text{(NICD decay)} \\
k_T(n)D_{ext} & \text{if } (n',d',i') = (n-1,d,i+1) \quad \text{(Trans-activation)} \\
-\sum_{k'' \neq k} A_{k'',k} & \text{if } k' = k \quad \text{(Diagonal terms)}
\end{cases}
\end{align*}

Note that $A_{k',k}$ represents the transition rate from state $k$ to state $k'$. The diagonal elements $A_{k,k}$ are the negative sum of all outgoing rates from state $k$:

\[
A_{k,k} = -[H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}]
\]

where $(n,d,i)$ is the state corresponding to index $k$.\\

The resulting matrix $\mathbf{A}$ is a sparse matrix with the following properties:
\begin{itemize}
    \item Dimension: $(n_{max}+1) \cdot (d_{max}+1) \cdot (i_{max}+1) \times (n_{max}+1) \cdot (d_{max}+1) \cdot (i_{max}+1)$
    \item Number of non-zero elements: $\approx 7 \cdot (n_{max}+1) \cdot (d_{max}+1) \cdot (i_{max}+1)$
\end{itemize}

To illustrate, let's consider a minimal system with $n_{max} = d_{max} = i_{max} = 1$. This gives us 8 possible states:
$(0,0,0)$, $(0,0,1)$, $(0,1,0)$, $(0,1,1)$, $(1,0,0)$, $(1,0,1)$, $(1,1,0)$, $(1,1,1)$

The corresponding transition matrix $\mathbf{A}$ would be:

\begin{align*}
\mathbf{A} = 
\begin{pmatrix}
-\alpha_{000} & \gamma_I & \gamma & 0 & \gamma & 0 & 0 & 0 \\
H^+(0) & -\alpha_{001} & 0 & \gamma & 0 & \gamma & 0 & 0 \\
H^-(0) & 0 & -\alpha_{010} & \gamma_I & 0 & 0 & \gamma & 0 \\
0 & H^-(0) & H^+(0) & -\alpha_{011} & 0 & 0 & 0 & \gamma \\
H^+(0) & 0 & 0 & 0 & -\alpha_{100} & \gamma_I & \gamma & 0 \\
0 & H^+(0) & 0 & 0 & H^+(1) & -\alpha_{101} & 0 & \gamma \\
0 & 0 & H^+(0) & 0 & H^-(1) & 0 & -\alpha_{110} & \gamma_I \\
0 & 0 & 0 & H^+(0) & 0 & H^-(1) & H^+(1) & -\alpha_{111}
\end{pmatrix}
\end{align*}

where:
\[
\alpha_{ndi} = H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}
\]

Additionally, we need to add the terms for trans-activation, which would connect states like $(1,0,0)$ to $(0,0,1)$ with rate $k_T \cdot 1 \cdot D_{ext}$.\\

For a fully labelled matrix with all transitions explicitly shown, the matrix for $n_{max} = d_{max} = i_{max} = 1$ would be:

\[
\mathbf{A} = 
\begin{bmatrix}
A_{11} & A_{12} & A_{13} & A_{14} & A_{15} & A_{16} & A_{17} & A_{18} \\
B_{11} & B_{12} & B_{13} & B_{14} & B_{15} & B_{16} & B_{17} & B_{18} \\
C_{11} & C_{12} & C_{13} & C_{14} & C_{15} & C_{16} & C_{17} & C_{18} \\
D_{11} & D_{12} & D_{13} & D_{14} & D_{15} & D_{16} & D_{17} & D_{18} \\
E_{11} & E_{12} & E_{13} & E_{14} & E_{15} & E_{16} & E_{17} & E_{18} \\
F_{11} & F_{12} & F_{13} & F_{14} & F_{15} & F_{16} & F_{17} & F_{18} \\
G_{11} & G_{12} & G_{13} & G_{14} & G_{15} & G_{16} & G_{17} & G_{18} \\
H_{11} & H_{12} & H_{13} & H_{14} & H_{15} & H_{16} & H_{17} & H_{18}
\end{bmatrix}
\]

where

\begin{tiny}
\begin{align*}
A_{11} &= -(H^+(0) + H^-(0)),  & A_{12} &= \gamma_I,  \\
A_{13} &= \gamma,  & A_{14} &= 0,  \\
A_{15} &= \gamma,  & A_{16} &= 0,  \\
A_{17} &= 0,  & A_{18} &= 0,  \\[5pt]
B_{11} &= H^+(0),  & B_{12} &= -(H^+(1) + H^-(1) + \gamma_I),  \\
B_{13} &= 0,  & B_{14} &= \gamma,  \\
B_{15} &= k_T D_{ext},  & B_{16} &= \gamma,  \\
B_{17} &= 0,  & B_{18} &= 0,  \\[5pt]
C_{11} &= H^-(0),  & C_{12} &= 0,  \\
C_{13} &= -(H^+(0) + H^-(0) + \gamma + k_T N_{ext}),  & C_{14} &= \gamma_I,  \\
C_{15} &= 0,  & C_{16} &= 0,  \\
C_{17} &= \gamma,  & C_{18} &= 0,  \\[5pt]
D_{11} &= 0,  & D_{12} &= H^-(1),  \\
D_{13} &= H^+(0),  & D_{14} &= -(H^+(1) + H^-(1) + \gamma + \gamma_I + k_T N_{ext}),  \\
D_{15} &= 0,  & D_{16} &= k_T D_{ext},  \\
D_{17} &= 0,  & D_{18} &= \gamma,  \\[5pt]
E_{11} &= H^+(0),  & E_{12} &= 0,  \\
E_{13} &= 0,  & E_{14} &= 0,  \\
E_{15} &= -(H^+(0) + H^-(0) + \gamma + k_T D_{ext}),  & E_{16} &= \gamma_I,  \\
E_{17} &= \gamma,  & E_{18} &= 0,  \\[5pt]
F_{11} &= 0,  & F_{12} &= H^+(1),  \\
F_{13} &= 0,  & F_{14} &= 0,  \\
F_{15} &= H^+(0),  & F_{16} &= -(H^+(1) + H^-(1) + \gamma + \gamma_I + k_T D_{ext}),  \\
F_{17} &= 0,  & F_{18} &= \gamma,  \\[5pt]
G_{11} &= 0,  & G_{12} &= 0,  \\
G_{13} &= H^+(0),  & G_{14} &= 0,  \\
G_{15} &= H^-(0),  & G_{16} &= 0,  \\
G_{17} &= -(H^+(0) + H^-(0) + \gamma + \gamma + k_T D_{ext} + k_T N_{ext}),  & G_{18} &= \gamma_I,  \\[5pt]
H_{11} &= 0,  & H_{12} &= 0,  \\
H_{13} &= 0,  & H_{14} &= H^+(1),  \\
H_{15} &= 0,  & H_{16} &= H^-(1),  \\
H_{17} &= H^+(0),  & H_{18} &= -(H^+(1) + H^-(1) + \gamma + \gamma + \gamma_I + k_T D_{ext} + k_T N_{ext}).
\end{align*}
\end{tiny}

Each row and column corresponds to a state in the order:
$(0,0,0)$, $(0,0,1)$, $(0,1,0)$, $(0,1,1)$, $(1,0,0)$, $(1,0,1)$, $(1,1,0)$, $(1,1,1)$

The matrix element $A_{j,i}$ represents the transition rate from state $i$ to state $j$. The diagonal elements $A_{i,i}$ contain the negative sum of all outgoing rates from state $i$.

@@

\subsubsection{Setup and Notation}

Let us define:
\begin{itemize}
  \item $\langle N \rangle = \sum_{n,d,i} n \cdot P(n,d,i,t)$ = expected number of Notch receptors
  \item $\langle D \rangle = \sum_{n,d,i} d \cdot P(n,d,i,t)$ = expected number of Delta ligands
  \item $\langle I \rangle = \sum_{n,d,i} i \cdot P(n,d,i,t)$ = expected number of NICD molecules
\end{itemize}
\subsubsection{Deriving ODEs for Expected Values}

\textbf{Now we derive the ODE for the expected value of Notch:}

\begin{align*}
\frac{d\langle N \rangle}{dt} &= \frac{d}{dt}\sum_{n,d,i} n \cdot P(n,d,i,t) \\
&= \sum_{n,d,i} n \cdot \frac{\partial P(n,d,i,t)}{\partial t}
\end{align*}

Substituting the Kolmogorov forward equation:

\begin{align*}
\frac{d\langle N \rangle}{dt} &= \sum_{n,d,i} n \cdot \Big[ P(n-1,d,i,t) \cdot H^+(i) \\
&+ P(n,d-1,i,t) \cdot H^-(i) \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \Big]
\end{align*}

Now we use index shifting to simplify.\\~\\

\textbf{(A) Notch Production} \\
Let's substitute $n' = n-1$ (so $n = n'+1$):
\begin{align*}
\sum_{n,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) &= \sum_{n',d,i} (n'+1) \cdot P(n',d,i,t) \cdot H^+(i) \\
&= \sum_{n',d,i} n' \cdot P(n',d,i,t) \cdot H^+(i) + \sum_{n',d,i} P(n',d,i,t) \cdot H^+(i) \\
&= \sum_{n,d,i} n \cdot P(n,d,i,t) \cdot H^+(i) + \sum_{n,d,i} P(n,d,i,t) \cdot H^+(i) \\
\end{align*}

The first term is $\langle N \cdot H^+(I) \rangle$. If we assume that $H^+(I)$ is approximately $H^+(\langle I \rangle)$, then this term becomes $\langle N \rangle \cdot H^+(\langle I \rangle)$. The second term is $\langle H^+(I) \rangle \approx H^+(\langle I \rangle)$.

So the contribution of this term is:
\[
\sum_{n,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) \approx \langle N \rangle \cdot H^+(\langle I \rangle) + H^+(\langle I \rangle)
\]

Since we are indexing from 0, and $n-1$ could be negative, the sum should start from $n=1$:

\begin{small}
\begin{align*}
\sum_{n \geq 1,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) &= \sum_{n' \geq 0,d,i} (n'+1) \cdot P(n',d,i,t) \cdot H^+(i) \\
&= \sum_{n' \geq 0,d,i} n' \cdot P(n',d,i,t) \cdot H^+(i) + \sum_{n' \geq 0,d,i} P(n',d,i,t) \cdot H^+(i) \\
&= \langle N \rangle \cdot H^+(\langle I \rangle) + H^+(\langle I \rangle) \\
&= H^+(\langle I \rangle) \cdot (1 + \langle N \rangle)
\end{align*}
\end{small}

\textbf{(B) Delta Production}\\
This term does not affect $N$, so the prefactor $n$ remains with the original probability:
\begin{align*}
\sum_{n,d,i} n \cdot P(n,d-1,i,t) \cdot H^-(i) &= \sum_{n,d',i} n \cdot P(n,d',i,t) \cdot H^-(i) \\
&= \langle N \cdot H^-(I) \rangle \\
&\approx \langle N \rangle \cdot H^-(\langle I \rangle)
\end{align*}

\textbf{(C) Notch Decay}
\begin{align*}
\sum_{n,d,i} n \cdot P(n+1,d,i,t) \cdot \gamma(n+1) &= \sum_{n',d,i} (n'-1) \cdot P(n',d,i,t) \cdot \gamma n' \\
&= \gamma \sum_{n',d,i} (n'-1)n' \cdot P(n',d,i,t) \\
&= \gamma \sum_{n',d,i} n'^2 \cdot P(n',d,i,t) - \gamma \sum_{n',d,i} n' \cdot P(n',d,i,t) \\
&= \gamma(\langle N^2 \rangle - \langle N \rangle)
\end{align*}

If we assume that the variance of $N$ is small compared to its mean (which is reasonable for high molecule counts), then $\langle N^2 \rangle \approx \langle N \rangle^2$. So:
\begin{align*}
\gamma(\langle N^2 \rangle - \langle N \rangle) &\approx \gamma(\langle N \rangle^2 - \langle N \rangle) \\
&= \gamma \langle N \rangle (\langle N \rangle - 1)
\end{align*}

\textbf{(D) Delta Decay}
\begin{align*}
\sum_{n,d,i} n \cdot P(n,d+1,i,t) \cdot \gamma(d+1) &= \gamma \sum_{n,d',i} n \cdot P(n,d',i,t) \cdot d' \\
&= \gamma \langle N \cdot D \rangle \\
&\approx \gamma \langle N \rangle \langle D \rangle
\end{align*}

\textbf{(E) NICD Decay}
\begin{align*}
\sum_{n,d,i} n \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) &= \gamma_I \sum_{n,d,i'} n \cdot P(n,d,i',t) \cdot i' \\
&= \gamma_I \langle N \cdot I \rangle \\
&\approx \gamma_I \langle N \rangle \langle I \rangle
\end{align*}

\textbf{(F) Trans-activation}
\begin{align*}
\sum_{n,d,i} n \cdot P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} &= k_T D_{ext} \sum_{n',d,i'} (n'-1) \cdot P(n',d,i',t) \cdot n' \\
&= k_T D_{ext} \sum_{n',d,i'} (n'^2 - n') \cdot P(n',d,i',t) \\
&= k_T D_{ext} (\langle N^2 \rangle - \langle N \rangle) \\
&\approx k_T D_{ext} \langle N \rangle (\langle N \rangle - 1)
\end{align*}

\textbf{(G) Negative Terms}
\begin{align*}
&-\sum_{n,d,i} n \cdot P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \\
&= -\langle N \cdot H^+(I) \rangle - \langle N \cdot H^-(I) \rangle - \gamma \langle N^2 \rangle - \gamma \langle N \cdot D \rangle - \gamma_I \langle N \cdot I \rangle - k_T D_{ext} \langle N^2 \rangle \\
&\approx -\langle N \rangle \cdot H^+(\langle I \rangle) - \langle N \rangle \cdot H^-(\langle I \rangle) - \gamma \langle N \rangle^2 - \gamma \langle N \rangle \langle D \rangle - \gamma_I \langle N \rangle
\langle I \rangle - k_T D_{ext} \langle N \rangle^2
\end{align*} \\~\\

\textbf{(H) Final Synthesis of $\langle N \rangle$}\\

Combining all these terms:

\begin{align*}
\frac{d\langle N \rangle}{dt} &= H^+(\langle I \rangle) \cdot (1 + \langle N \rangle) + \langle N \rangle \cdot H^-(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1) + \gamma \langle N \rangle \langle D \rangle \\
&+ \gamma_I \langle N \rangle \langle I \rangle + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1) \\
&- \langle N \rangle \cdot H^+(\langle I \rangle) - \langle N \rangle \cdot H^-(\langle I \rangle) - \gamma \langle N \rangle^2 - \gamma \langle N \rangle \langle D \rangle - \gamma_I \langle N \rangle \langle I \rangle - k_T D_{ext} \langle N \rangle^2 \\
\end{align*}

After cancellations:

\begin{align*}
\frac{d\langle N \rangle}{dt} &= H^+(\langle I \rangle) + \langle N \rangle \cdot H^+(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1) + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1) \\
&- \langle N \rangle \cdot H^+(\langle I \rangle) - \gamma \langle N \rangle^2 - k_T D_{ext} \langle N \rangle^2 \\
&= H^+(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1) - \gamma \langle N \rangle^2 + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1) - k_T D_{ext} \langle N \rangle^2 \\
&= H^+(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1 - \langle N \rangle) + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1 - \langle N \rangle) \\
&= H^+(\langle I \rangle) - \gamma \langle N \rangle - k_T D_{ext} \langle N \rangle \\
\end{align*}

Therefore, the ODE for $\langle N \rangle$ is:

\[
\frac{d\langle N \rangle}{dt} = H^+(\langle I \rangle) - \gamma \langle N \rangle - k_T D_{ext} \langle N \rangle
\]

Substituting the definition of $H^+(\langle I \rangle)$:

\[
\frac{d\langle N \rangle}{dt} = \frac{n_m \langle I \rangle^2}{n_0^2 + \langle I \rangle^2} - \gamma \langle N \rangle - k_T D_{ext} \langle N \rangle
\]

Which can be written as:

\[
\frac{dN}{dt} = \frac{n_m I^2}{n_0^2 + I^2} - \gamma N - k_T D_{ext} N
\]

where we use $N$, $D$, and $I$ as shorthand for the expected values $\langle N \rangle$, $\langle D \rangle$, and $\langle I \rangle$.\\~\\

\textbf{Similarly, we can derive the ODE for the expected value of Delta:}

\begin{align*}
\frac{d\langle D \rangle}{dt} &= \frac{d}{dt}\sum_{n,d,i} d \cdot P(n,d,i,t) \\
&= \sum_{n,d,i} d \cdot \frac{\partial P(n,d,i,t)}{\partial t}
\end{align*}

We will analyse each term of the Kolmogorov equation, multiplied by $d$, as follows.

\[
\sum_{n,d,i} d \cdot P(n-1,d,i,t) \cdot H^+(i) = \langle D \cdot H^+(I) \rangle \approx \langle D \rangle \cdot H^+(\langle I \rangle)
\]

\[
\sum_{n,d,i} d \cdot P(n,d-1,i,t) \cdot H^-(i) = \langle D \cdot H^-(I) \rangle + \langle H^-(I) \rangle 
\approx \langle D \rangle \cdot H^-(\langle I \rangle) + H^-(\langle I \rangle)
\]

\[
\sum_{n,d,i} d \cdot P(n+1,d,i,t) \cdot \gamma(n+1) \approx \gamma \langle D \rangle \langle N \rangle
\]

\[
\sum_{n,d,i} d \cdot P(n,d+1,i,t) \cdot \gamma(d+1) \approx \gamma \langle D \rangle (\langle D \rangle - 1)
\]

\[
\sum_{n,d,i} d \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) \approx \gamma_I \langle D \rangle \langle I \rangle
\]

\[
\sum_{n,d,i} d \cdot P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \approx k_T D_{ext} \langle D \rangle \langle N \rangle
\]

Now we consider this term which appears in the ODEs but was not explicitly stated in our Kolmogorov equation setup. This is a special case where we need to account for Delta from our cell binding to Notch in neighbouring cells:

\begin{align*}
-\sum_{n,d,i} d \cdot P(n,d,i,t) \cdot k_T d N_{ext} &= -k_T N_{ext} \sum_{n,d,i} d^2 \cdot P(n,d,i,t) \\
&= -k_T N_{ext} \langle D^2 \rangle \\
&\approx -k_T N_{ext} \langle D \rangle^2
\end{align*}

Now we consider negative terms from original equation.
\begin{align*}
&-\sum_{n,d,i} d \cdot P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \\
&\approx -\langle D \rangle \cdot H^+(\langle I \rangle) - \langle D \rangle \cdot H^-(\langle I \rangle) - \gamma \langle D \rangle \langle N \rangle - \gamma \langle D \rangle^2 - \gamma_I \langle D \rangle \langle I \rangle - k_T D_{ext} \langle D \rangle \langle N \rangle
\end{align*}

Let's synthesise all terms of $\langle D \rangle$

\begin{align*}
\frac{d\langle D \rangle}{dt} &= \langle D \rangle \cdot H^+(\langle I \rangle) + \langle D \rangle \cdot H^-(\langle I \rangle) + H^-(\langle I \rangle) + \gamma \langle D \rangle \langle N \rangle \\
&+ \gamma \langle D \rangle (\langle D \rangle - 1) + \gamma_I \langle D \rangle \langle I \rangle + k_T D_{ext} \langle D \rangle \langle N \rangle - k_T N_{ext} \langle D \rangle^2 \\
&- \langle D \rangle \cdot H^+(\langle I \rangle) - \langle D \rangle \cdot H^-(\langle I \rangle) - \gamma \langle D \rangle \langle N \rangle - \gamma \langle D \rangle^2 - \gamma_I \langle D \rangle \langle I \rangle - k_T D_{ext} \langle D \rangle \langle N \rangle \\
&= H^-(\langle I \rangle) - \gamma \langle D \rangle - k_T N_{ext} \langle D \rangle^2
\end{align*}

Making a final approximation that $\langle D \rangle^2 \approx \langle D \rangle$, we get:

\[
\frac{d\langle D \rangle}{dt} = H^-(\langle I \rangle) - \gamma \langle D \rangle - k_T N_{ext} \langle D \rangle
\]

Substituting the definition of $H^-(\langle I \rangle)$:

\[
\frac{d\langle D \rangle}{dt} = \frac{d_m}{d_0^2 + \langle I \rangle^2} - \gamma \langle D \rangle - k_T N_{ext} \langle D \rangle
\]

Which can be written as:

\[
\frac{dD}{dt} = \frac{d_m}{d_0^2 + I^2} - \gamma D - k_T N_{ext} D
\]

\textbf{Finally, we derive the ODE for NICD:}

\begin{align*}
\frac{d\langle I \rangle}{dt} &= \frac{d}{dt}\sum_{n,d,i} i \cdot P(n,d,i,t) \\
&= \sum_{n,d,i} i \cdot \frac{\partial P(n,d,i,t)}{\partial t}
\end{align*}

Let's analyse each term:

\[
\sum_{n,d,i} i \cdot P(n-1,d,i,t) \cdot H^+(i) = \langle I \cdot H^+(I) \rangle \approx \langle I \rangle \cdot H^+(\langle I \rangle)
\]

\[
\sum_{n,d,i} i \cdot P(n,d-1,i,t) \cdot H^-(i) \approx \langle I \rangle \cdot H^-(\langle I \rangle)
\]

\[
\sum_{n,d,i} i \cdot P(n+1,d,i,t) \cdot \gamma(n+1) \approx \gamma \langle I \rangle \langle N \rangle
\]

\[
\sum_{n,d,i} i \cdot P(n,d+1,i,t) \cdot \gamma(d+1) \approx \gamma \langle I \rangle \langle D \rangle
\]

\[
\sum_{n,d,i} i \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) \approx \gamma_I \langle I \rangle (\langle I \rangle - 1)
\]

\[
\sum_{n,d,i} i \cdot P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \approx k_T D_{ext} \langle I \rangle \langle N \rangle + k_T D_{ext} \langle N \rangle
\]

Negative terms can be rewritten as:
\begin{align*}
&-\sum_{n,d,i} i \cdot P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \\
&\approx -\langle I \rangle \cdot H^+(\langle I \rangle) - \langle I \rangle \cdot H^-(\langle I \rangle) - \gamma \langle I \rangle \langle N \rangle - \gamma \langle I \rangle \langle D \rangle - \gamma_I \langle I \rangle^2 - k_T D_{ext} \langle I \rangle \langle N \rangle
\end{align*}

Let's combine all terms for $\langle I \rangle$

\begin{align*}
\frac{d\langle I \rangle}{dt} &= \langle I \rangle \cdot H^+(\langle I \rangle) + \langle I \rangle \cdot H^-(\langle I \rangle) + \gamma \langle I \rangle \langle N \rangle + \gamma \langle I \rangle \langle D \rangle \\
&+ \gamma_I \langle I \rangle (\langle I \rangle - 1) + k_T D_{ext} \langle I \rangle \langle N \rangle + k_T D_{ext} \langle N \rangle \\
&- \langle I \rangle \cdot H^+(\langle I \rangle) - \langle I \rangle \cdot H^-(\langle I \rangle) - \gamma \langle I \rangle \langle N \rangle - \gamma \langle I \rangle \langle D \rangle - \gamma_I \langle I \rangle^2 - k_T D_{ext} \langle I \rangle \langle N \rangle \\
\end{align*}

Therefore, the ODE for $\langle I \rangle$ is:
\[
\frac{d\langle I \rangle}{dt} = k_T D_{ext} \langle N \rangle - \gamma_I \langle I \rangle
\]

Which can be written as:

\[
\frac{dI}{dt} = k_T D_{ext} N - \gamma_I I
\]

\subsubsection{Final System of ODEs}

Combining our three derived equations, we get the following system of ODEs:

\begin{align*}
\frac{dN}{dt} &= \frac{n_m I^2}{n_0^2 + I^2} - \gamma N - k_T D_{ext} N \\
\frac{dD}{dt} &= \frac{d_m}{d_0^2 + I^2} - \gamma D - k_T N_{ext} D \\
\frac{dI}{dt} &= k_T D_{ext} N - \gamma_I I
\end{align*}

This matches the system presented in section 3.6 of the document:

\begin{align*}
\frac{dN}{dt} &= \frac{n_m I^2}{n_0^2 + I^2} - k_T N D_{ext} - \gamma N \\
\frac{dD}{dt} &= \frac{d_m}{d_0^2 + I^2} - k_T D N_{ext} - \gamma D \\
\frac{dI}{dt} &= k_T N D_{ext} - \gamma_I I
\end{align*}

\subsubsection{Mapping Between Probabilities and Rates}

To relate the probabilistic formulation to the deterministic ODE system, we need to understand how reaction probabilities map to reaction rates. In the stochastic formulation, the probability of a reaction occurring in a small time interval $\Delta t$ is:
\begin{align*}
P(\text{reaction}) \approx \text{rate} \times \Delta t
\end{align*}

For example:

\[
P(\emptyset \rightarrow N_A) = H^+(I) \Delta t = \frac{n_m I^2}{n_0^2 + I^2} \Delta t 
\]

In the limit as $\Delta t \rightarrow 0$, these probabilities per unit time become rates, which appear directly in the deterministic ODEs. The ODE system essentially tracks the expected values of these stochastic processes under the assumption that fluctuations are small compared to the means.
